---
title: "Predictive measures in measures.R"
output:
  html_document:
    toc: true
    math_method: mathjax
---

```{r root, include = FALSE}
# Paths in this file are relative to the pipeline folder.
knitr::opts_knit$set(root.dir = "..")
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("workflow/scripts/lib/dgp.R")
source("workflow/scripts/lib/glm.R")
source("workflow/scripts/lib/measures.R")
```

This file states every predictive measure of
`workflow/scripts/lib/measures.R`. Then it computes each measure on example
data of each family, and compares model A with model B. Equation numbers
refer to `Predictive_measures_and_model_comparison_in_the_loo_package.pdf`.

`measures.R` has two layers:

1. **Pointwise scores.** `pointwise_scores_family()` evaluates the predictive
   distribution $p(\tilde y_i)$ of each observation at $y_i$. This is the
   only layer that depends on the family, because each family has its own
   predictive distribution.
2. **Measures.** `MEASURE_SPEC` builds every measure from one score. This
   layer does not depend on the family. It follows `.measure_spec` of the loo
   package, branch `pred_measure`.

# Pointwise scores

| Score | Pointwise value | Orientation | Families |
| :-- | :-- | :-- | :-- |
| `elpd` | $\log p(y_i)$ (2.16) | utility | all |
| `sqe` | $(y_i - \mathrm{E}[\tilde y_i])^2$ | loss | all |
| `ae` | $\lvert y_i - \mathrm{E}[\tilde y_i] \rvert$ | loss | all |
| `rps` | $\mathrm{E}\lvert X - y_i \rvert - \tfrac12 \mathrm{E}\lvert X - X' \rvert$ (2.20) | loss | all |
| `srps` | $-\dfrac{\mathrm{E}\lvert X - y_i \rvert}{\mathrm{E}\lvert X - X' \rvert} - \tfrac12 \log \mathrm{E}\lvert X - X' \rvert$ (2.21) | utility | all |
| `acc` | $\mathbb 1[\hat y_i = y_i]$, $\hat y_i = \mathbb 1[p(\tilde y_i = 1) \ge 0.5]$ | utility | binomial |

$X$ and $X'$ are independent draws from $p(\tilde y_i)$. For a binary
outcome, $\mathrm{E}[\tilde y_i] = p(\tilde y_i = 1)$, so `sqe` is the Brier
score (2.26).

# Measures

## Definition for one model

$n$ is the number of observations and $s_i$ is the pointwise score.
$\mathrm{MSE}(y) = \frac1n \sum_i (y_i - \bar y)^2$.

| Measure | Score | Estimate | Loss |
| :-- | :-- | :-- | :-- |
| `elpd` | `elpd` | $\sum_i s_i$ | no |
| `mlpd` | `elpd` | $\frac1n \sum_i s_i$ | no |
| `ic` | `elpd` | $-2 \sum_i s_i$ | yes |
| `mse` | `sqe` | $\frac1n \sum_i s_i$ | yes |
| `rmse` | `sqe` | $\sqrt{\frac1n \sum_i s_i}$ | yes |
| `r2` | `sqe` | $1 - \frac1n \sum_i s_i \,/\, \mathrm{MSE}(y)$ | no |
| `brier` | `sqe` | $\frac1n \sum_i s_i$ | yes |
| `mae` | `ae` | $\frac1n \sum_i s_i$ | yes |
| `rps` | `rps` | $\frac1n \sum_i s_i$ | yes |
| `srps` | `srps` | $\frac1n \sum_i s_i$ | no |
| `acc` | `acc` | $\frac1n \sum_i s_i$ | no |
| `bacc` | `acc` | $\frac12 \sum_{c} \frac1{n_c} \sum_{i \in c} s_i$ | no |

`brier`, `acc` and `bacc` need a binary outcome. `measures_for_family()`
drops them for the other families.

## Difference of two models

`loo_difference()` gives $\Delta = U(M_a) - U(M_b)$ on the utility scale. A
loss is negated first, so $\Delta > 0$ always favours $M_a$. The
`diff_method` of `MEASURE_SPEC` selects the SE:

```{r spec}
spec <- data.frame(
  measure = names(MEASURE_SPEC),
  score = MEASURE_SCORE,
  loss = vapply(MEASURE_SPEC, `[[`, logical(1), "loss"),
  diff_method = vapply(MEASURE_SPEC, `[[`, character(1), "diff_method"),
  row.names = NULL
)
knitr::kable(spec)
```

- **`sum` and `mean`.** Let $d_i$ be the pointwise utility difference from
  `pointwise_terms()`. Then $\Delta = \sum_i d_i$ with
  $\widehat{\mathrm{SE}} = \sqrt n \, \mathrm{sd}(d_i)$ (2.12, 2.14), or
  $\Delta = \bar d$ with $\widehat{\mathrm{SE}} = \mathrm{sd}(d_i) / \sqrt n$
  (2.11, 2.13).
- **`rmse`** (3.3), `se_diff_rmse()`. Let
  $z_i = s_{a,i} / \mathrm{RMSE}_a - s_{b,i} / \mathrm{RMSE}_b$. Then
  $\widehat{\mathrm{SE}} = \tfrac12 \sqrt{\mathrm{Var}(z_i) / n}$.
- **`r2`** (3.4, 3.5), `se_diff_r2()`. Let $\delta_i = s_{a,i} - s_{b,i}$
  and $c = \bar\delta / \mathrm{MSE}(y)$. Then
  $\widehat{\mathrm{SE}} = \sqrt{\mathrm{Var}(\delta_i - c\,(y_i - \bar y)^2) / n} \,/\, \mathrm{MSE}(y)$.
- **`bacc`**, `se_diff_bacc()`. The classes are disjoint, so their
  variances add:
  $\widehat{\mathrm{SE}} = \tfrac12 \sqrt{\sum_c \mathrm{Var}(d_i : i \in c) / n_c}$.

The pipeline differs from loo in degenerate cases. An SE of zero, or a class
with fewer than two observations, gives NA, so that the trial is dropped.
loo reports an SE of zero.

# Example data

Each family uses one training set of $n = 64$ observations, no outlier and
$\beta_t = 1$. Model A drops the last covariate. Model B uses all covariates,
so B is the true model. A pooled test set gives the target, the
out-of-sample value of each measure.

```{r example-setup}
N_OBS <- 64
# The test sets of config/small.yaml. The binomial and poisson scores use
# quadrature, so these families use fewer test sets.
N_SETS_TEST <- c(gaussian = 200, binomial = 50, poisson = 20)
BETA <- make_beta_coeff(beta_t = 1)

# LOO scores of the training set and scores of the test set, for model A
# and model B. Training sets first, then test sets, as rule generate_data.
example_family <- function(family, seed = 5) {
  set.seed(seed)
  train <- make_data_family(family, 1, N_OBS, N_OBS, BETA, out_dev = 0)
  test <- pool_sets(make_data_family(family, N_SETS_TEST[[family]], N_OBS,
                                     N_OBS, BETA, out_dev = 0))
  X <- train$X[, , 1]
  y <- train$y[, 1]
  measures <- measures_for_family(names(MEASURE_SPEC), family)
  scores <- scores_for_family(SCORES_BINOMIAL, family)
  models <- lapply(c(A = "A", B = "B"), function(m) {
    cols <- model_columns(m, ncol(X))
    X_m <- X[, cols, drop = FALSE]
    fit <- fit_model_family(family, X_m, y)
    loo_pred <- loo_predictive_family(family, X_m, y, start = fit$beta_hat)
    test_pred <- predict_new_family(family, fit,
                                    test$X[, cols, drop = FALSE])
    list(loo = pointwise_scores_family(family, y, loo_pred, scores),
         test = pointwise_scores_family(family, test$y, test_pred, scores))
  })
  list(family = family, y = y, y_test = test$y, measures = measures,
       models = models)
}

examples <- lapply(c(gaussian = "gaussian", binomial = "binomial",
                     poisson = "poisson"), example_family)
```

## The outcomes

```{r outcomes}
outcome_summary <- function(ex) {
  data.frame(family = ex$family, n = length(ex$y), mean = mean(ex$y),
             sd = sd(ex$y), min = min(ex$y), max = max(ex$y),
             n_test = length(ex$y_test))
}
knitr::kable(do.call(rbind, lapply(examples, outcome_summary)),
             digits = 2, row.names = FALSE)
```

## The pointwise LOO scores

The first five observations of model B:

```{r pointwise, results = "asis"}
for (ex in examples) {
  s <- as.data.frame(ex$models$B$loo)
  print(knitr::kable(head(cbind(y = ex$y, s), 5), digits = 3,
                     caption = ex$family))
  cat("\n")
}
```

## Each measure for each model

`measure_estimate()` gives a measure of one model on its own scale, from
`MEASURE_SPEC`. A sum is scaled to $n$ observations, so that the test value
and the LOO value are comparable.

```{r measure-estimate}
measure_estimate <- function(measure, scores, y, n = length(y)) {
  spec <- MEASURE_SPEC[[measure]]
  s <- scores[[spec$score]]
  if (spec$diff_method == "measure_specific") return(spec$estimate(s, y))
  weight <- if (is.null(spec$weight)) 1 else spec$weight
  m <- mean(weight * s)
  if (spec$diff_method == "sum") n * m else m
}

model_table <- function(ex) {
  one <- function(m, set) {
    y <- if (set == "loo") ex$y else ex$y_test
    vapply(ex$measures, measure_estimate, numeric(1),
           scores = ex$models[[m]][[set]], y = y, n = length(ex$y))
  }
  data.frame(measure = ex$measures,
             loo_A = one("A", "loo"), test_A = one("A", "test"),
             loo_B = one("B", "loo"), test_B = one("B", "test"),
             row.names = NULL)
}
```

```{r model-tables, results = "asis"}
for (ex in examples) {
  print(knitr::kable(model_table(ex), digits = 3, caption = ex$family))
  cat("\n")
}
```

## Comparison of model A with model B

`loo_difference()` gives the LOO estimate of $\Delta = U(A) - U(B)$ and its
SE. `target_difference()` gives the same difference on the test set. B is
the true model, so the target is negative for every measure. $z$ is the
estimate divided by its SE.

```{r compare-table}
compare_table <- function(ex) {
  a <- ex$models$A
  b <- ex$models$B
  test_mean <- function(m) lapply(m$test, mean)
  by_class <- function(m) {
    lapply(m$test, function(s) vapply(split(s, ex$y_test), mean, numeric(1)))
  }
  est <- vapply(ex$measures, loo_difference, numeric(2),
                loo_a = a$loo, loo_b = b$loo, y = ex$y)
  tgt <- vapply(ex$measures, target_difference, numeric(1),
                test_a = test_mean(a), test_b = test_mean(b),
                n_obs = length(ex$y),
                y_test_var = mean((ex$y_test - mean(ex$y_test))^2),
                by_class_a = by_class(a), by_class_b = by_class(b))
  data.frame(measure = ex$measures, estimate = est[1, ], se = est[2, ],
             z = est[1, ] / est[2, ], target = tgt, row.names = NULL)
}
```

```{r compare-tables, results = "asis"}
for (ex in examples) {
  print(knitr::kable(compare_table(ex), digits = 3, caption = ex$family))
  cat("\n")
}
```

## Check against the single-model values

For every measure, the LOO difference equals the difference of the
single-model values on the utility scale. The largest absolute deviation
over all families and measures:

```{r consistency}
deviation <- function(ex) {
  mt <- model_table(ex)
  sign <- ifelse(vapply(MEASURE_SPEC[ex$measures], `[[`, logical(1), "loss"),
                 -1, 1)
  est <- vapply(ex$measures, function(m) {
    loo_difference(m, ex$models$A$loo, ex$models$B$loo, ex$y)[1]
  }, numeric(1))
  abs(est - sign * (mt$loo_A - mt$loo_B))
}
max(unlist(lapply(examples, deviation)))
```

## Check of rule compare_models

The chunks run `workflow/scripts/compare_models.R` on the examples, with a
mock `snakemake` object. Each family is one cell with two chunks. Both
chunks hold the example training set, as trial 1 and trial 2. The rows of
the script must equal `compare_table()` for each trial.

```{r compare-models-input}
N_CHUNKS <- 2
dir <- tempfile("compare_models_")
dir.create(dir)
write_rds <- function(x, name) {
  path <- file.path(dir, name)
  saveRDS(x, path)
  path
}

# The output of rule compute_measure for one model and one chunk.
score_file <- function(ex, m, trial) {
  s <- ex$models[[m]]
  by_class <- if (ex$family == "binomial") {
    lapply(s$test, function(v) {
      as.matrix(vapply(split(v, ex$y_test), mean, numeric(1)))
    })
  }
  write_rds(list(trials = trial, loo = lapply(s$loo, as.matrix),
                 test = lapply(s$test, mean), test_by_class = by_class),
            sprintf("%s_%s_%d.rds", ex$family, m, trial))
}
# The score files in the order of grid_files(): cells, then chunks.
score_files <- function(m) {
  as.list(unlist(lapply(examples, function(ex) {
    lapply(seq_len(N_CHUNKS), function(k) score_file(ex, m, k))
  })))
}

cells <- lapply(examples, function(ex) {
  list(family = ex$family, n_obs = N_OBS, beta_t = 1, out_dev = 0,
       n_chunks = N_CHUNKS)
})
input <- list(
  a = score_files("A"),
  b = score_files("B"),
  train = lapply(examples, function(ex) {
    write_rds(list(y = matrix(ex$y, N_OBS, N_CHUNKS)),
              paste0(ex$family, "_train.rds"))
  }),
  info = lapply(examples, function(ex) {
    write_rds(list(y_test_var = mean((ex$y_test - mean(ex$y_test))^2)),
              paste0(ex$family, "_info.rds"))
  })
)
```

```{r compare-models-run}
setClass("MockSnakemake", representation(
  input = "list", output = "list", params = "list", log = "list",
  source = "function"
))
env <- new.env()
env$snakemake <- new(
  "MockSnakemake",
  input = input,
  output = list(file.path(dir, "trials.rds")),
  params = list(cells = cells, measures = as.list(names(MEASURE_SPEC))),
  log = list(file.path(dir, "compare_models.log")),
  source = function(f) {
    sys.source(file.path("workflow/scripts", f), envir = env)
    # Keep the log connection of start_log(), to close it after the run.
    if (f == "lib/log.R") {
      start_log <- env$start_log
      env$start_log <- function(path) env$log_con <- start_log(path)
    }
  }
)
sys.source("workflow/scripts/compare_models.R", envir = env)
# start_log() sends the messages to the log file. Send them back, and
# close the log file.
sink(type = "message")
close(env$log_con)
trials <- readRDS(file.path(dir, "trials.rds"))
```

The output has `r nrow(trials)` rows. The first rows:

```{r compare-models-head}
knitr::kable(head(trials, 8), digits = 3)
```

Rows of each cell and trial:

```{r compare-models-count}
table(cell = trials$cell, trial = trials$trial)
```

The largest absolute deviation from `compare_table()`, and the check that
the missing values are at the same places:

```{r compare-models-check}
expected <- do.call(rbind, lapply(examples, function(ex) {
  ct <- compare_table(ex)
  do.call(rbind, lapply(seq_len(N_CHUNKS), function(k) {
    data.frame(trial = k, ct[c("measure", "estimate", "se", "target")],
               cell = ex$family)
  }))
}))
stopifnot(
  identical(trials$cell, expected$cell),
  identical(as.integer(trials$trial), expected$trial),
  identical(trials$measure, expected$measure)
)
cols <- c("estimate", "se", "target")
got <- unname(as.matrix(trials[cols]))
want <- unname(as.matrix(expected[cols]))
max(abs(got - want), na.rm = TRUE)
identical(is.na(got), is.na(want))
```
