---
title: "Heavy-tailed covariates: the t scenario"
output:
  html_document:
    toc: true
    toc_float: true
params:
  # Results folder and cell table, relative to the pipeline folder.
  results: results/paper
  cells: config/cells_paper.tsv
  config: config/paper.yaml
---

```{r root, include = FALSE}
# Paths in this file are relative to the pipeline folder.
knitr::opts_knit$set(root.dir = "..")
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r data}
cells <- utils::read.delim(params$cells)
cfg <- yaml::read_yaml(params$config)
x_dfs <- sort(unique(cells$x_df), decreasing = TRUE)
x_dfs_t <- x_dfs[is.finite(x_dfs)]
```

This note adds one factor to the paper grid of Sivula et al. (2025): the
distribution of the covariates. In the paper the covariates are standard
normal. Here they are also standard t, with degrees of freedom from
`r max(x_dfs_t)` down to 1 (Cauchy). Everything else stays as in the paper:
the gaussian family, the measure elpd, and the cells of Fig. 5 and Fig. 6.

Render it from the pipeline folder:

```sh
Rscript -e 'rmarkdown::render("notes/xdf_scenario.Rmd", output_dir = ".")'
```

# Data generating process

`make_data_gaussian()` in `workflow/scripts/lib/dgp.R` draws one data set
of $n$ observations as

$$
y_i = \beta_0 + \beta_1 x_{i1} + \beta_t x_{i2} + \varepsilon_i,
\qquad
\varepsilon_i \sim \mathrm{N}(\mu_i, \sigma^2),
\qquad
\beta = (0, 1, \beta_t), \quad \sigma^2 = 1.
$$

Model A omits $x_{2}$. Model B includes it. Both are normal linear
regressions with the exact LOO predictive. The quantity of interest is the
elpd difference $\mathrm{elpd}_A - \mathrm{elpd}_B$.

## Covariates

The covariates $x_{ij}$ are independent. With $\nu$ = `x_df` they are

$$
x_{ij} = \frac{z_{ij}}{\sqrt{w_{ij} / \nu}},
\qquad
z_{ij} \sim \mathrm{N}(0, 1),
\qquad
w_{ij} = F^{-1}_{\chi^2_\nu}(u_{ij}),
\quad u_{ij} \sim \mathrm{U}(0, 1),
$$

so $x_{ij} \sim t_\nu$. For $\nu = \infty$ the denominator is 1, and
$x_{ij} = z_{ij}$ is the normal covariate of the paper.

- The t is not rescaled to unit variance. Its variance is
  $\nu / (\nu - 2)$, which is infinite for $\nu \le 2$.
- $z$ and $\varepsilon$ are drawn first, as in the paper. $u$ is drawn last,
  and only for a finite $\nu$. So the normal cells keep the random numbers
  of the paper run.
- All t cells share the same $z$, $\varepsilon$ and $u$. Two cells that
  differ only in $\nu$ differ only in the tails of the covariates. This
  gives common random numbers across $\nu$.

```{r densities, fig.width = 7, fig.height = 3.2, fig.cap = "Density of the covariate, linear and log scale."}
x <- seq(-8, 8, length.out = 801)
cols <- grDevices::hcl.colors(length(x_dfs), "viridis")
dens <- sapply(x_dfs, function(df) if (is.finite(df)) dt(x, df) else dnorm(x))
lab <- ifelse(is.finite(x_dfs), paste0("t(", x_dfs, ")"), "normal")
par(mfrow = c(1, 2), mar = c(4, 4, 1, 1), mgp = c(2.2, 0.6, 0), bty = "l")
matplot(x, dens, type = "l", lty = 1, col = cols, xlim = c(-5, 5),
        xlab = "x", ylab = "density")
legend("topright", legend = lab, col = cols, lty = 1, bty = "n", cex = 0.7)
matplot(x, dens, type = "l", lty = 1, col = cols, log = "y",
        xlab = "x", ylab = "density (log)")
```

## Outlier

As in the paper, the first observation gets the mean shift

$$
\mu_1 = \Delta \sqrt{\sigma^2 + \lVert \beta \rVert^2},
\qquad \mu_i = 0 \text{ for } i > 1,
$$

with $\Delta$ = `out_dev`. The scale $\sqrt{\sigma^2 + \lVert\beta\rVert^2}$
is the SD of $y$ for a normal covariate. It is kept for the t covariate, so
the outlier has the same size in every cell.

## Test sets

The test sets come from the same DGP as the training sets, with the same
$\nu$. The target is the elpd difference on
`r cfg$n_sets_test_gaussian` pooled test sets of size $n$.

# Simulation scenarios

```{r grid}
fmt <- function(v) paste(sort(unique(v)), collapse = ", ")
knitr::kable(data.frame(
  factor = c("n", "beta_t", "out_dev", "x_df"),
  levels = c(fmt(cells$n_obs), fmt(cells$beta_t), fmt(cells$out_dev),
             paste(c(fmt(x_dfs_t), "Inf (normal)"), collapse = ", "))
), col.names = c("Factor", "Levels"))
```

- The full product of the factors gives `r nrow(cells)` cells.
  `r sum(is.finite(cells$x_df))` of them have a t covariate.
- Each cell has `r cfg$n_sets_train` trials (training sets).
- A t cell has the name of its normal cell plus `_t<df>`, for example
  `n32_b0_o0_t1`.

The figures for the normal covariate keep their names,
`{fig}_gaussian_elpd.pdf`. The figures for a t covariate are
`{fig}_gaussian_elpd_t<df>.pdf`. Each figure shows one $\nu$.

# Results

Each section shows one paper figure. Each tab shows one covariate
distribution, from the normal to the Cauchy.

```{r png}
# The figures are PDFs. pdftoppm turns the first page into a PNG.
png_dir <- file.path(params$results, "figs", "png")
dir.create(png_dir, showWarnings = FALSE, recursive = TRUE)
to_png <- function(fig, df) {
  stem <- paste0(fig, "_gaussian_elpd", if (is.finite(df)) paste0("_t", df))
  pdf <- file.path(params$results, "figs", paste0(stem, ".pdf"))
  png <- file.path(png_dir, stem)
  if (!file.exists(pdf)) return(NA_character_)
  system2("pdftoppm", c("-png", "-r", "110", "-singlefile", shQuote(pdf),
                        shQuote(png)))
  # Absolute, because pandoc resolves paths from the notes folder.
  normalizePath(paste0(png, ".png"))
}

tabs <- function(fig) {
  for (df in x_dfs) {
    cat("###", if (is.finite(df)) sprintf("t(%g)", df) else "normal", "\n\n")
    path <- to_png(fig, df)
    if (is.na(path)) cat("Figure not found.\n\n") else
      cat(sprintf("![](%s){width=100%%}\n\n", path))
  }
}
```

## Calibration {.tabset}

PIT histograms of the normal approximation, as Fig. 6 of the paper (normal
column). One column per $n$. For each $\beta_t$ one row without and one
with the outlier. A calibrated SE gives a flat histogram inside the 99 %
band.

```{r calibration, results = "asis"}
tabs("calibration")
```

## Joint distribution {.tabset}

The LOO estimate against the target, without the outlier, as Fig. 5 of the
paper. Points on the green diagonal are estimated without error.

```{r joint, results = "asis"}
tabs("joint")
```

## Moments {.tabset}

Mean/SD and skewness of the estimate, the target and the error against
$n$, without the outlier, as Fig. 9 of the paper. Bands are 95 %
Bayesian-bootstrap intervals.

```{r moments, results = "asis"}
tabs("moments")
```

## SE ratio {.tabset}

The stated SE divided by the actual SD of the LOO error, without the
outlier, as `simulated/plot_all__var_ratioi_n_b.py`. A ratio below 1 means
that the SE is too small.

```{r var_ratio, results = "asis"}
tabs("var_ratio")
```

## Error {.tabset}

The LOO error in units of its own SD, as `simulated/plot_all__err_n_b_both.py`.
The short line is the mean.

```{r err, results = "asis"}
tabs("err")
```

## Error direction {.tabset}

The error times the sign of the target, as
`simulated/plot_all__errdirection_n_b_both.py`. A positive value means that
the estimate favours the better model too strongly.

```{r errdirection, results = "asis"}
tabs("errdirection")
```
