---
title: "Non-normal covariates: heavy tails and skew"
output:
  html_document:
    toc: true
    toc_float: true
params:
  # Results folder and cell table, relative to the pipeline folder.
  results: results/paper
  cells: config/cells_paper.tsv
  config: config/paper.yaml
---

```{r root, include = FALSE}
# Paths in this file are relative to the pipeline folder.
knitr::opts_knit$set(root.dir = "..")
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r data}
cells <- utils::read.delim(params$cells)
cfg <- yaml::read_yaml(params$config)
# parse_x_dist() and SKEWT_ALPHA come from the simulation code.
dgp <- new.env()
sys.source("workflow/scripts/lib/dgp.R", dgp)
parse_x_dist <- dgp$parse_x_dist
skewt_alpha <- dgp$SKEWT_ALPHA

# Order: normal, t from many to few degrees of freedom, skew-t, gamma.
x_dists <- unique(cells$x_dist)
x_pars <- lapply(x_dists, parse_x_dist)
x_dists <- x_dists[order(
  match(sapply(x_pars, `[[`, "dist"), c("normal", "t", "skewt", "gamma")),
  -sapply(x_pars, `[[`, "par")
)]
x_kind <- sapply(x_dists, function(d) parse_x_dist(d)$dist)
x_par <- sapply(x_dists, function(d) parse_x_dist(d)$par)
x_dfs_t <- sort(x_par[x_kind == "t"], decreasing = TRUE)

# Label of a covariate distribution, for example "skew-t(5)".
x_label <- function(x_dist) {
  d <- parse_x_dist(x_dist)
  switch(d$dist,
    normal = "normal",
    t = sprintf("t(%g)", d$par),
    skewt = sprintf("skew-t(%g)", d$par),
    gamma = sprintf("gamma(%g)", d$par)
  )
}
x_labels <- sapply(x_dists, x_label)
```

This note adds one factor to the paper grid of Sivula et al. (2025): the
distribution of the covariates. In the paper the covariates are standard
normal. Here they also have two other kinds of distribution:

- **Heavy tails:** standard t, with degrees of freedom from
  `r max(x_dfs_t)` down to 1 (Cauchy).
- **Skew:** `r paste(x_labels[x_kind %in% c("skewt", "gamma")], collapse = " and ")`,
  standardised to mean 0 and variance 1. Their fourth moment is finite.
  So they show the effect of skew without infinite moments.

Everything else stays as in the paper: the gaussian family, the measure
elpd, and the cells of Fig. 5 and Fig. 6.

Render it from the pipeline folder:

```sh
Rscript -e 'rmarkdown::render("notes/xdf_scenario.Rmd", output_dir = ".")'
```

# Data generating process

`make_data_gaussian()` in `workflow/scripts/lib/dgp.R` draws one data set
of $n$ observations as

$$
y_i = \beta_0 + \beta_1 x_{i1} + \beta_t x_{i2} + \varepsilon_i,
\qquad
\varepsilon_i \sim \mathrm{N}(\mu_i, \sigma^2),
\qquad
\beta = (0, 1, \beta_t), \quad \sigma^2 = 1.
$$

Model A omits $x_{2}$. Model B includes it. Both are normal linear
regressions with the exact LOO predictive. The quantity of interest is the
elpd difference $\mathrm{elpd}_A - \mathrm{elpd}_B$.

## Covariates

The covariates $x_{ij}$ are independent. `x_dist` in the cell table sets
their distribution: `normal`, `t<df>`, `skewt<df>` or `gamma<k>`.

### t

With $\nu$ degrees of freedom the covariates are

$$
x_{ij} = \frac{z_{ij}}{\sqrt{w_{ij} / \nu}},
\qquad
z_{ij} \sim \mathrm{N}(0, 1),
\qquad
w_{ij} = F^{-1}_{\chi^2_\nu}(u_{ij}),
\quad u_{ij} \sim \mathrm{U}(0, 1),
$$

so $x_{ij} \sim t_\nu$. For $\nu = \infty$ the denominator is 1, and
$x_{ij} = z_{ij}$ is the normal covariate of the paper.

- The t is not rescaled to unit variance. Its variance is
  $\nu / (\nu - 2)$, which is infinite for $\nu \le 2$.

### Skew-t

The Azzalini skew-t with $\nu$ degrees of freedom and slant
$\alpha$ = `r skewt_alpha` is

$$
s_{ij} = \frac{\delta\,\lvert z^0_{ij}\rvert + \sqrt{1 - \delta^2}\, z_{ij}}
  {\sqrt{w_{ij} / \nu}},
\qquad
\delta = \frac{\alpha}{\sqrt{1 + \alpha^2}},
\qquad
z^0_{ij} \sim \mathrm{N}(0, 1),
$$

with $z_{ij}$ and $w_{ij}$ as for the t. The covariate is
$x_{ij} = (s_{ij} - m) / \sqrt{v}$, with the mean $m$ and the variance $v$
of $s_{ij}$. It has mean 0 and variance 1. A positive $\alpha$ gives a
long right tail. The variance needs $\nu > 2$. The fourth moment needs
$\nu > 4$.

### Gamma

With shape $k$ the covariates are

$$
x_{ij} = \frac{g_{ij} - k}{\sqrt{k}},
\qquad
g_{ij} = F^{-1}_{\mathrm{Gamma}(k, 1)}(u_{ij}).
$$

They have mean 0 and variance 1. All moments are finite. The support is
$x > -\sqrt{k}$, so the left tail is short and the right tail is
long.

### Random numbers

- $z$ and $\varepsilon$ are drawn first, as in the paper. $u$ is drawn
  next, and only for a covariate other than the normal. The skew-t draws
  $z^0$ last. So the normal cells keep the random numbers of the paper
  run.
- All non-normal cells share the same $z$, $\varepsilon$ and $u$. Two t
  cells that differ only in $\nu$ differ only in the tails of the
  covariates. This gives common random numbers across the covariate
  distributions.
- The gamma covariate uses $u$ only, not $z$. So it is a monotone
  function of the same $u$ that sets the tails of the t.

```{r skewt-moments}
# Raw moments 1 to 4 of the skew-t s before the standardisation. s is a
# skew-normal divided by sqrt(w / nu); the two parts are independent.
skewt_raw <- function(nu, alpha) {
  delta <- alpha / sqrt(1 + alpha^2)
  b <- sqrt(2 / pi)
  sn <- c(b * delta, 1, b * delta * (3 - delta^2), 3)
  k <- 1:4
  scale <- ifelse(k < nu,
    (nu / 2)^(k / 2) * exp(lgamma((nu - k) / 2) - lgamma(nu / 2)), Inf)
  sn * scale
}

# Mean, variance, skewness and kurtosis of the skew-t s.
skewt_moments <- function(nu, alpha) {
  m <- skewt_raw(nu, alpha)
  v <- m[2] - m[1]^2
  c3 <- m[3] - 3 * m[1] * m[2] + 2 * m[1]^3
  c4 <- m[4] - 4 * m[1] * m[3] + 6 * m[1]^2 * m[2] - 3 * m[1]^4
  c(mean = m[1], var = v, skew = c3 / v^1.5, kurt = c4 / v^2)
}
```

```{r densities, fig.width = 7, fig.height = 3.2, fig.cap = "Density of the covariate, linear and log scale. Dashed: the skewed covariates."}
dens_x <- function(x, x_dist) {
  d <- parse_x_dist(x_dist)
  k <- d$par
  out <- switch(d$dist,
    normal = dnorm(x),
    t = dt(x, k),
    skewt = {
      m <- skewt_moments(k, skewt_alpha)
      s <- m[["mean"]] + sqrt(m[["var"]]) * x
      sqrt(m[["var"]]) * 2 * dt(s, k) *
        pt(skewt_alpha * s * sqrt((k + 1) / (k + s^2)), k + 1)
    },
    gamma = sqrt(k) * dgamma(k + sqrt(k) * x, k)
  )
  # NA, not 0, outside the support, for the log scale.
  replace(out, out == 0, NA)
}
x <- seq(-8, 8, length.out = 801)
cols <- grDevices::hcl.colors(length(x_dists), "viridis")
ltys <- ifelse(x_kind %in% c("skewt", "gamma"), 2, 1)
dens <- sapply(x_dists, function(d) dens_x(x, d))
par(mfrow = c(1, 2), mar = c(4, 4, 1, 1), mgp = c(2.2, 0.6, 0), bty = "l")
matplot(x, dens, type = "l", lty = ltys, col = cols, xlim = c(-5, 5),
        xlab = "x", ylab = "density")
legend("topright", legend = x_labels, col = cols, lty = ltys, bty = "n",
       cex = 0.7)
matplot(x, dens, type = "l", lty = ltys, col = cols, log = "y",
        xlab = "x", ylab = "density (log)")
```

## Moments of the covariate

A moment of order $k$ of $t_\nu$ is finite only for $k < \nu$. The elpd
difference depends on $x_2$ and on $x_2^2$ (see below), so four moments
matter. For $t_\nu$ they are:

| Quantity | Finite for | Value |
| :-- | :-- | :-- |
| $\mathrm{E}[x_2]$ | $\nu > 1$ | 0 |
| $\mathrm{Var}[x_2] = \mathrm{E}[x_2^2]$ | $\nu > 2$ | $\nu / (\nu - 2)$ |
| Skewness of $x_2$ | $\nu > 3$ | 0 |
| $\mathrm{Var}[x_2^2]$ | $\nu > 4$ | $\dfrac{3\nu^2}{(\nu - 2)(\nu - 4)} - \left(\dfrac{\nu}{\nu - 2}\right)^2$ |

The skew-t and the gamma have mean 0 and variance 1. So
$\mathrm{Var}[x_2^2] = \kappa - 1$, with the kurtosis $\kappa$. The gamma
with shape $k$ has the skewness $2 / \sqrt{k}$ and
$\kappa = 3 + 6 / k$. The skew-t moments come from the moments of the
skew-normal and of $(w / \nu)^{-1/2}$.

```{r moments-table}
fmt3 <- function(v) format(v, digits = 3)
mom_x <- function(x_dist) {
  d <- parse_x_dist(x_dist)
  k <- d$par
  switch(d$dist,
    normal = c("0", "1", "0", "2"),
    t = c(
      if (k > 1) "0" else "undefined",
      fmt3(if (k > 2) k / (k - 2) else Inf),
      if (k > 3) "0" else "undefined",
      fmt3(if (k > 4) 3 * k^2 / ((k - 2) * (k - 4)) - (k / (k - 2))^2 else Inf)
    ),
    skewt = {
      m <- skewt_moments(k, skewt_alpha)
      c("0", "1", fmt3(m[["skew"]]), fmt3(m[["kurt"]] - 1))
    },
    gamma = c("0", "1", fmt3(2 / sqrt(k)), fmt3(2 + 6 / k))
  )
}
tab <- t(sapply(x_dists, mom_x))
knitr::kable(data.frame(covariate = x_labels, tab),
  col.names = c("Covariate", "E[x2]", "Var[x2]", "Skewness", "Var[x2^2]"),
  row.names = FALSE)
```

Why $x_2^2$ matters: model A omits $\beta_t x_2$, so this term is part of
its residual. The pointwise log predictive density of model A is then about

$$
\log p_A(y_i) \approx -\tfrac{1}{2}\log(2\pi\hat\sigma_A^2)
  - \frac{(\beta_t x_{i2} + \varepsilon_i)^2}{2\hat\sigma_A^2},
$$

and the pointwise elpd difference contains $\beta_t^2 x_{i2}^2$ and
$\beta_t x_{i2} \varepsilon_i$. The normal approximation sums these terms
and uses $\widehat{\mathrm{SE}} = \sqrt{n}\,\mathrm{sd}(\text{pointwise})$.
This relies on the central limit theorem, so it needs a finite variance of
the pointwise terms, that is $\nu > 4$.

The expected pattern, for $\beta_t > 0$:

- **$\nu \ge 5$:** all three moments are finite. The covariate behaves
  like the normal one, only with more spread.
- **$\nu = 3$:** $\mathrm{Var}[x_2]$ is finite, but
  $\mathrm{Var}[x_2^2]$ is infinite. A few points with large $|x_2|$
  dominate the sum. The sample SD underestimates the spread, so the SE is
  too small. The effect is moderate, because $\mathrm{E}[x_2^2]$ is still
  finite.
- **$\nu = 1$ (Cauchy):** $\mathrm{E}[x_2]$ is undefined and
  $\mathrm{E}[x_2^2]$ is infinite. Often one point dominates the sum. A
  larger $n$ does not help, because a larger sample has a more extreme
  maximum.
- **Skew-t and gamma:** the mean and the variance are as for the normal,
  and $\mathrm{Var}[x_2^2]$ is finite. So the central limit theorem holds.
  But $x_2$ has a long right tail, so $x_2^2$ and the pointwise terms are
  strongly skewed. For a small $n$ the sum is then still skewed, and the
  normal approximation can be poorly calibrated. A larger $n$ helps.
  The skew-t has the larger $\mathrm{Var}[x_2^2]$, so the effect is
  stronger for the skew-t than for the gamma.

For $\beta_t = 0$ model A misses nothing, so $x_2^2$ is not in the
residual. But a point with a large $|x_2|$ has a leverage near 1 in model
B, and its LOO term in model B is then extreme.

The outlier shift $\Delta \sqrt{\sigma^2 + \lVert\beta\rVert^2}$ does not
depend on the covariate distribution. For a t covariate with a small $\nu$ the spread of $y$ is much larger than for
the normal covariate, so the same shift is a weaker outlier.

These explanations are hypotheses. The pointwise terms per trial were not
analysed.

## Outlier

As in the paper, the first observation gets the mean shift

$$
\mu_1 = \Delta \sqrt{\sigma^2 + \lVert \beta \rVert^2},
\qquad \mu_i = 0 \text{ for } i > 1,
$$

with $\Delta$ = `out_dev`. The scale $\sqrt{\sigma^2 + \lVert\beta\rVert^2}$
is the SD of $y$ for a normal covariate. It is kept for the other
covariates, so the outlier has the same size in every cell. The skew-t and
the gamma have variance 1, so for them the scale is again the SD of $y$.

## Test sets

The test sets come from the same DGP as the training sets, with the same
covariate distribution. The target is the elpd difference on
`r cfg$n_sets_test_gaussian` pooled test sets of size $n$.

# Simulation scenarios

```{r grid}
fmt <- function(v) paste(sort(unique(v)), collapse = ", ")
knitr::kable(data.frame(
  factor = c("n", "beta_t", "out_dev", "x_dist"),
  levels = c(fmt(cells$n_obs), fmt(cells$beta_t), fmt(cells$out_dev),
             paste(x_dists, collapse = ", "))
), col.names = c("Factor", "Levels"))
```

- The full product of the factors gives `r nrow(cells)` cells.
  `r sum(cells$x_dist != "normal")` of them have a covariate other than
  the normal.
- Each cell has `r cfg$n_sets_train` trials (training sets).
- A non-normal cell has the name of its normal cell plus `_<x_dist>`,
  for example `n32_b0_o0_t1` or `n32_b0_o0_skewt5`.

The figures for the normal covariate keep their names,
`{fig}_gaussian_elpd.pdf`. The figures for the other covariates are
`{fig}_gaussian_elpd_<x_dist>.pdf`. Each figure shows one covariate
distribution.

# Results

Each section shows one paper figure. Each tab shows one covariate
distribution: the normal, the t from many degrees of freedom to the Cauchy,
then the skewed covariates.

```{r png}
# The figures are PDFs. pdftoppm turns the first page into a PNG.
png_dir <- file.path(params$results, "figs", "png")
dir.create(png_dir, showWarnings = FALSE, recursive = TRUE)
to_png <- function(fig, x_dist) {
  stem <- paste0(fig, "_gaussian_elpd",
                 if (x_dist != "normal") paste0("_", x_dist))
  pdf <- file.path(params$results, "figs", paste0(stem, ".pdf"))
  png <- file.path(png_dir, stem)
  if (!file.exists(pdf)) return(NA_character_)
  system2("pdftoppm", c("-png", "-r", "110", "-singlefile", shQuote(pdf),
                        shQuote(png)))
  # Absolute, because pandoc resolves paths from the notes folder.
  normalizePath(paste0(png, ".png"))
}

tabs <- function(fig) {
  for (x_dist in x_dists) {
    cat("###", x_label(x_dist), "\n\n")
    path <- to_png(fig, x_dist)
    if (is.na(path)) cat("Figure not found.\n\n") else
      cat(sprintf("![](%s){width=100%%}\n\n", path))
  }
}
```

## Calibration {.tabset}

PIT histograms of the normal approximation, as Fig. 6 of the paper (normal
column). One column per $n$. For each $\beta_t$ one row without and one
with the outlier. A calibrated SE gives a flat histogram inside the 99 %
band.

```{r calibration, results = "asis"}
tabs("calibration")
```

## Joint distribution {.tabset}

The LOO estimate against the target, without the outlier, as Fig. 5 of the
paper. Points on the green diagonal are estimated without error.

```{r joint, results = "asis"}
tabs("joint")
```

## Moments {.tabset}

Mean/SD and skewness of the estimate, the target and the error against
$n$, without the outlier, as Fig. 9 of the paper. Bands are 95 %
Bayesian-bootstrap intervals.

```{r moments, results = "asis"}
tabs("moments")
```

## SE ratio {.tabset}

The stated SE divided by the actual SD of the LOO error, without the
outlier, as `simulated/plot_all__var_ratioi_n_b.py`. A ratio below 1 means
that the SE is too small.

```{r var_ratio, results = "asis"}
tabs("var_ratio")
```

## Error {.tabset}

The LOO error in units of its own SD, as `simulated/plot_all__err_n_b_both.py`.
The short line is the mean.

```{r err, results = "asis"}
tabs("err")
```

## Error direction {.tabset}

The error times the sign of the target, as
`simulated/plot_all__errdirection_n_b_both.py`. A positive value means that
the estimate favours the better model too strongly.

```{r errdirection, results = "asis"}
tabs("errdirection")
```
